Final Project

### 2024-04-22 Draft_report.Rmd
added limiatations, intro and refrneces

**Introduction**

**Motivation/relevance of the study**
Global ocean temperature is on the rise which has significant implications for 
this dynamic and productive biome. Among the many services the ocean provides, 
primary production (of oxygen) is among the most important. In fact, 
approximately 50% of primary production of oxygen comes from marine 
phytoplankton. It has been proven (largely through work by Duke University 
Marine Lab's Dr. Zacakry Johnson and Dr. Dana Hunt) that these microbial 
systems change significantly with seasonal changes (which include dominantly temperature changes, but also isolation and day length changes). These changes 
alter the way that the ocean cycles nutrients, stores carbon, and produces 
oxygen. Hence, temperature is an important variable in the consideration of 
much ocean modeling. This highlights the importance of accurate temperature 
prediction. Understanding what future climate and temperature looks like 
allows for better prediction of what other ocean cycles will look like.

@@

The following data comes from a long running time series study out at Piver's 
Island Coastal Observatory (PICO) which is located at the Duke University 
Marine Lab. The sytudy monitors the ambient conditions such as turbidity, pH, temperature, and salinity. For this study, only temperature was able to be 
included. The temperature is reported as mean monthly temperature.

@@

```{r, echo=FALSE}
#We important the monthly temperature data
file_path <- "/home/guest/TSA_Janka/WaterTemperature_TSA_FinalProject/Monthly Data_Temperature2011-2021.xlsx"
temperature_data <- read_excel(file_path, sheet = "Sheet1")
```

@@

```{r, echo=FALSE}
#Upon initial examination, the data appears to be strongly seasonal with very little trend.
plot(temperature_ts, main = "Original dataset")
```
The temperature shows cyclical behavior with peaks and troughs that correspond 
to expected seasonal variations. The amplitude of these oscillations appears 
consistent over the years, indicating a stable seasonal effect without 
significant year-over-year changes in peak or trough temperatures.
From a visual inspection, there are no apparent outliers or disruptions in the 
seasonal pattern, suggesting that the dataset is clean and well-maintained. 
The regularity of the pattern would likely make it suitable for forecasting 
using seasonal models, such as SARIMA or TBATS, which could exploit the 
periodicity inherent in the data.
```{r, echo=FALSE}
#The ACF and Pacf tell us that there is ovbious and strong seasonality and some auto-regressive component up to about 6 lags.
P1 <-Acf(temperature_ts, main = "ACF on original data")
```
The graph illustrates the Autocorrelation Function (ACF) applied to the original temperature data set. Autocorrelation is a mathematical representation of the 
degree of similarity between a given time series and a lagged version of itself 
over successive time intervals. The ACF is plotted against various lag values, 
which span from 0 to 24.
The y-axis represents the autocorrelation coefficient, ranging from -0.5 to 0.5, 
while the x-axis represents the lag in terms of the number of time units. Each 
vertical bar corresponds to the autocorrelation coefficient at a specific lag.
Notably, the graph shows a pattern of spikes at regular intervals, which 
suggests a seasonal pattern in the data. The presence of these spikes at 
consistent intervals can be indicative of the seasonality in the dataset, which correlates with the seasonal fluctuations seen in the time series plot of 
the original dataset.
The blue dashed lines represent the significance bounds. Any spike that extends 
beyond these bounds is considered statistically significant. In this ACF plot, 
several lags have autocorrelation values that cross the significance threshold,
confirming that the data exhibit a strong seasonal component at these lags.

```{r, echo=FALSE}}
P2 <- Pacf(temperature_ts, main = "PACF on original data")
```
In this PACF plot, most of the spikes are within the significance bounds, 
suggesting that most of the autocorrelations in the original data can be 
accounted for by the immediate preceding values.

@@

print(adf.test(temperature_ts, alternative = "stationary"))
```
The Augmented Dickey-Fuller Test result indicates a Dickey-Fuller statistic of 
-11.153 with a lag order of 5, and a p-value of 0.01, suggesting that the null 
hypothesis of a unit root can be rejected and the time series is stationary.
```{r, echo=FALSE}
#confirming that timeseries transformation works as expected.
#str(temperature_ts)
```

@@

```{r, echo=FALSE}
#Confirms Training and Test sets were correctly made
#start(train_ts)
#end(train_ts)
#start(test_ts)
#end(test_ts)
```

@@

*Create Score table*
```{r, echo=FALSE}
scores <- as.data.frame(
  rbind(Sarima_scores, ArimaFour_scores, stlf_scores, TBATS_scores, SSES_scores, NN_scores)
  )
row.names(scores) <- c("Sarima", "ARIMA+Fourier","ETS+STL","TBATS","SSES", "NN")

best_model_index <- which.min(scores[,"RMSE"])
cat("The best model by RMSE is:", row.names(scores[best_model_index,]))                       
```

>>>>

**Introduction**
Marine microbe account for 50% of marine primary production and dominate the 
biomass of the ocean. These organism are vital for life on earth and ocean 
health but are at serious risk in high temperatures worsened by climate change. 
The first step in supporting these ecological communities and hence mitigating 
the damages done by climate change is to understand what future climate 
scenarios look like. Predictions for future climate and temperature can varry by 
up to 10 degrees (if not more). With so much uncertainty surrounding climate 
future, it is of the utmost importance to have as much acurate modeling for 
future climate scenarios and hence our group hopes to make reasonable forecasts 
of surface water temperature at the Duke University Marine Lab research facility.

**Motivation/relevance of the study**
Global ocean temperature is on the rise which has significant implications for 
this dynamic and productive biome. Among the many services the ocean provides, 
primary production (of oxygen) is among the most important. In fact, 
approximately 50% of primary production of oxygen comes from marine 
phytoplankton. It has been proven (largely through work by Duke University 
Marine Lab's Dr. Zacakry Johnson and Dr. Dana Hunt) that these microbial 
systems change significantly with seasonal changes (which include dominantly 
temperature changes, but also isolation and day length changes). These changes 
alter the way that the ocean cycles nutrients, stores carbon, and produces 
oxygen. Hence, temperature is an important variable in the consideration of much 
ocean modeling. This highlights the importance of accurate temperature 
prediction. Understanding what future climate and temperature looks like allows 
for better prediction of what other ocean cycles will look like and how we may 
excpet the serives we recieve from the ocean to change with the climate.

@@

The following data comes from a long running time series study out at Piver's 
Island Coastal Observatory (PICO) which is located at the Duke University 
Marine Lab conducted by Drs. Zackary Johnson and Dana Hunt. The sytudy monitors 
the ambient conditions such as turbidity, pH, temperature, and salinity. For 
this study, only temperature was able to be included. The temperature is 
reported as mean monthly temperature.

@@

```{r, echo=FALSE}
#We important the monthly temperature data
file_path <-
  "~/WaterTemperature_TSA_FinalProject/Monthly Data_Temperature2011-2021.xlsx"
temperature_data <- read_excel(file_path, sheet = "Sheet1")
```

@@

```{r, echo=FALSE}
#Upon initial examination, the data appears to be seasonal with little trend.
plot(temperature_ts, main = "Original dataset")
```
The temperature shows cyclical behavior with peaks and troughs that correspond 
to expected seasonal variations. The amplitude of these oscillations appears 
consistent over the years, indicating a stable seasonal effect without 
significant year-over-year changes in peak or trough temperatures.
From a visual inspection, there are no apparent outliers or disruptions in the 
seasonal pattern, suggesting that the dataset is clean and well-maintained. 
The regularity of the pattern would likely make it suitable for forecasting 
using seasonal models, such as SARIMA or TBATS, which could exploit the 
periodicity inherent in the data.
```{r, echo=FALSE}
#The ACF and Pacf tell us that there is ovbious and strong seasonality and some auto-regressive component up to about 6 lags.
P1 <-Acf(temperature_ts, main = "ACF on original data")
```
The graph illustrates the Autocorrelation Function (ACF) applied to the original 
temperature data set. Autocorrelation is a mathematical representation of the 
degree of similarity between a given time series and a lagged version of itself 
over successive time intervals. The ACF is plotted against various lag values, 
which span from 0 to 24.
The y-axis represents the autocorrelation coefficient, ranging from -0.5 to 0.5, 
while the x-axis represents the lag in terms of the number of time units. Each 
vertical bar corresponds to the autocorrelation coefficient at a specific lag.
Notably, the graph shows a pattern of spikes at regular intervals, which 
suggests a seasonal pattern in the data. The presence of these spikes at 
consistent intervals can be indicative of the seasonality in the dataset, which 
correlates with the seasonal fluctuations seen in the time series plot of 
the original dataset.
The blue dashed lines represent the significance bounds. Any spike that extends 
beyond these bounds is considered statistically significant. In this ACF plot, 
several lags have autocorrelation values that cross the significance threshold,
confirming that the data exhibit a strong seasonal component at these lags.

```{r, echo=FALSE}
P2 <- Pacf(temperature_ts, main = "PACF on original data")
```
In this PACF plot, most of the spikes are within the significance bounds, 
suggesting that most of the autocorrelations in the original data can be 
accounted for by the immediate preceding values.

@@

print(adf.test(temperature_ts, alternative = "stationary"))
```
The Augmented Dickey-Fuller Test result indicates a Dickey-Fuller statistic of 
-11.153 with a lag order of 5, and a p-value of 0.01, suggesting that the null 
hypothesis of a unit root can be rejected and the time series is stationary.
```{r, echo=FALSE}
#confirming that timeseries transformation works as expected.
str(temperature_ts)
```

@@

```{r, echo=FALSE}
#Confirms Training and Test sets were correctly made
start(train_ts)
end(train_ts)
start(test_ts)
end(test_ts)
```

@@

*Create Score table*
```{r, echo=FALSE}
scores <- as.data.frame(
  rbind(
    Sarima_scores,
    ArimaFour_scores,
    stlf_scores,
    TBATS_scores,
    SSES_scores,
    NN_scores
  )
)
row.names(scores) <-
  c("Sarima", "ARIMA+Fourier", "ETS+STL", "TBATS", "SSES", "NN")

best_model_index <- which.min(scores[, "RMSE"])
cat("The best model by RMSE is:", row.names(scores[best_model_index, ]))                       
```

**Limitations**
Though we have produced accurate models and forecasts, there are a few limitations to be aware of. The first is regarding our prediction time frame We only have access to ~ 10 years of data which likely does not demonstrate enough of a trend to be recognized as the role of climate change in our models. We should expect increasing temepratures in the future as a result of climate change. Even if our model did capture this linear trend, the modeling would be based on the assumtions that the trend is constant (aka that emissions causing climate change are to remain exactly as they are now, neither increasing nor decreasing). This limits further the time frame for which projections and models are viable. Finally, the resolution of our data is not ideal; we have monthly aggregate data so we are projecting out monthly as well. This is not as helpful as a finer resolution may have been given the fact that monthy aggregate data does not accurately reflect the extreme temeprature spikes we may see on specific days. In the future, it would be usely to repeat this modeling ideally with daily data. Even geting multiple temepratures a day could had more complexity as we should except to see patterns of temeprature changes throughout the day.

**Selected Refrences**
Johnson, Z. I., Wheeler, B. J., Blinebry, S. K., Carlson, C. M., Ward, C., & Hunt, D. E. (2013a). Dramatic variability of the carbonate system at a temperate coastal ocean site (Beaufort, North Carolina, USA) is regulated by physical and biogeochemical processes on multiple timescales. PloS One, 8(12), e85117. https://doi.org/10.1371/journal.pone.0085117

McIntyre, A. D. (2010). Life in the world’s oceans. In Wiley eBooks. https://doi.org/10.1002/9781444325508

Wang, Z., Tsementzi, D., Williams, T. C., Juarez, D. L., Blinebry, S. K., Garcia, N. S., Sienkiewicz, B. K., Konstantinidis, K. T., Johnson, Z. I., & Hunt, D. E. (2020a). Environmental stability impacts the differential sensitivity of marine microbiomes to increases in temperature and acidity. The ISME Journal, 15(1), 19–28. https://doi.org/10.1038/s41396-020-00748-2
